Cultural Transmission Can Explain the Evolution of Cooperative Behavior

### 2020-07-06 ms/main.tex
reverted the last part and revised it from scratch

\textbf{Result 3: Oblique and vertical transmission of cooperation with symmetric horizontal transmission.}

Assuming equal horizontal transmission, i.e. $T_A = T_B = T$, and following the above procedure, we have
\begin{equation} \label{eq:equal_horizontal_transmission}
%\hat{p}' - \hat{p} =  \hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T))
\bar{w} (\hat{p}' - \hat{p}) =  \hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T))
\end{equation}

It is easy to see that $\hat{p}^*_0 =  0$ and $\hat{p}^*_1 = 1$ are equilibria of eq.~\ref{eq:equal_horizontal_transmission}, that is, both extinction and fixation of the cooperator phenotype are equilibria.

%TODO fix the following, you forgot wbar in eq:equal_horizontal_transmission
In order to do that we find the derivative  $f(\hat{p}) = \hat{p}'-\hat{p}$.
\begin{equation} \label{eq:derivative_of_phattag-phat}
 f'(\hat{p}) = 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
\end{equation}
We are interested in the case when when $\hat{p} = \hat{p}^*_1=1$ therefore
$f'(1) = -\alpha bvT + cv(1-T)$. Therefore, fixation of cooperation is stable when
\begin{equation} \label{eq:oblique_and_vertic_result2}
  \frac{b}{c}>\frac{1-T}{\alpha T}
\end{equation}
Which is the same condition we get in corollary 1.1.

@@

Next, we study the general case where $T_A \neq T_B$. We have
%TODO check again, I think you forgot wbar here, too
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \hat{p}'-\hat{p} = \beta_1 p^3 + \beta_2 p^2 + \beta_3 p
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{split}
  & \beta_1 = c(1-v)\cdot(T_A-T_B) -b (1-\alpha v) \cdot(T_A-T_B) \\
  & \beta_2 = (b(1-\alpha v)-c(1-v)-1)(T_A-T_B)-\alpha bv T_A + cv(1-T_B) \\
  & \beta_3 = \alpha bvT_A - cv(1-T_B) + (T_A-T_B)
\end{split}
\end{equation}

$f(\hat{p})$ is a cubic, and its three roots are the equilibria of the system, i.e. the solutions for $\hat{p}'=\hat{p}$.
As $\hat{p}^*_0 = 0 $ and $\hat{p}^*_1 = 1$ are equilibria, there is only one equilibrium left, $\hat{p}^*_2$, which solves $f(\hat{p})/\big(\hat{p}(1-\hat{p})\big)=0$. Solving this we find that
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^*_2 =  \frac{\alpha bvT_A + cvT_B + T_A-T_B-cv}{c(1-v)\cdot(T_A-T_B) -b (1-\alpha v) \cdot(T_A-T_B)}
\end{equation}

$\hat{p}^*_2$ is a legitimate equilibrium if $0<\hat{p}^*_2<1$.  
To determine if $\hat{p}^*_2$ is legitimate, we need the derivative of $f(\hat{p})$,\begin{equation} \label{eq:derivative_of_phattag-phat2}
  f'(\hat{p}) = 3\beta_0 \hat{p}^2 + 2\beta_1\hat{p}^2 + \beta_3.
 \end{equation}
First, we assume that $T_A>T_B$, and therefore $\beta_1<0$, because $c(1-v)-b(1-\alpha v)<0$ and $c<b$ $v<1$. % why?
In this case $f(\hat{p})$ goes to infinity when $\hat{p}$ goes to infinity. 
Therefore, if $\hat{p}^*_2$ is positive then $f'(0)= \beta_3$ must be negative, and
% TODO remove
%\begin{equation} \label{eq:derivative_of_phattag-phat_p=0}
%  f'(0) = \beta_3
% \end{equation} 
$\hat{p}_2^*>0$ when $\beta_3<0$. In addition, $\hat{p}^*_0=0$ is stable when $|f'(0)|<1$ therefore, when $-1<\beta_3<1$. We can show that 
\begin{equation} \label{eq:beta3-condition}
  \beta_3 = \alpha bvT_A - cv(1-T_B) + (T_A-T_B) > -cv(1-T_B) > -1
\end{equation}
Therefore, when $\hat{p}^*_2>0$ we get that $-1<\beta_3<0$ and $\hat{p}^*_0=0$ is locally stable.

Next we would like to find when $\hat{p}^*_2<1$.
If $T_A>T_B$ like before then the polynomial goes to minus infinity when $\hat{p}$ goes to infinity.
Hence, if $\hat{p}_2<1$ then $f'(1)$ must be negative
\begin{equation} \label{eq:derivative_of_phattag-phat_p=1}
  f'(0) = \beta_1 - \beta_3
\end{equation}
Therefore, $\hat{p}_2<1$ when $\beta_1-\beta_3<0$. In addition, $\hat{p}_1=1$ is stable when $|f'(1)|<1$ therefore, when $-1<\beta_1-\beta_3<1$. 
We can prove that when $0<\hat{p}_2<1$ hence $\beta_3<0$ and $\beta_1-\beta_3<0$ we get that $-1<\beta_1-\beta_3$:
\begin{equation} \label{eq:beta1-beta3-condition}
 \beta_1 - \beta_3 > - \beta_3 > 0 > -1
\end{equation}
which means that when $0<\hat{p}_2<1$ then $\hat{p}_1=1$ is always stable.

\textbf{result 2.3}: for the case when $T_A>T_B$ $\hat{p_2}$ is realistic when $\beta_1<\beta_3<0$ and is always unstable equilibrium.

\textbf{result 2.4}: for the case when $T_B>T_A$ $\hat{p_2}$ is realistic when $\beta_1>\beta_3>0$ and is always unstable equilibrium.

>>>>

\textbf{Result 3: Oblique and vertical transmission with symmetric horizontal transmission.}

For simplicity, assume that $T_A = T_B = T$ so we will get
\begin{equation} \label{eq:equal_horizontal_transmission}
% check this, I think you are missing a wbar on the left side, ie it should be
% \bar{\hat{p}' - \hat{p}} =  \hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T))
  \hat{p}' - \hat{p} =  \hat{p} ^2(-\alpha bvT + cv(1-T)) + \hat{p} (\alpha bvT -cv(1-T))
\end{equation}

We are looking for the equilibrium $\hat{p}^*$ which occurs when $\hat{p}' = \hat{p}$.
It is easy to verify that fixation of either phenotype is an equilibrium, that is,
\begin{equation} \label{eq:result2.1}
  \hat{p}^*_0 =  0, \, \hat{p}^*_1 = 1 .
\end{equation}
We wish to determine when fixation of cooperation is locally stable. % state the general condition on the derivative that implies local stability
To do that, we calculate the derivative of $f(\hat{p}) = \hat{p}'-\hat{p}$, 
\begin{equation} \label{eq:derivative_of_phattag-phat}
 f'(\hat{p}) = 2\cdot\hat{p} (-\alpha bvT + cv(1-T)) + (\alpha bvT -cv(1-T)).
\end{equation}
For $\hat{p} = \hat{p}^*_1=1$, the derivative is
$f'(1) = -\alpha bvT + cv(1-T)$, and therefore, fixation of cooperation is locally stable when
\begin{equation} \label{eq:oblique_and_vertic_result2}
  \frac{b}{c}>\frac{1-T}{\alpha T}
\end{equation}
Which is the same condition we get in corollary 1.1.

@@

We now study the general case where $T_A \neq T_B$. We have
\begin{equation} \label{eq:general_case_polynomial}% check, because I think wbar is missing
  f(\hat{p}) = \hat{p}'-\hat{p} = \beta_1p^3 + \beta_2 p^2 + \beta_3p + \beta_4
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{split}
  & \beta_1 = c(1-v)\cdot(T_A-T_B) -b (1-\alpha v) \cdot(T_A-T_B) \\
  & \beta_2 = (b(1-\alpha v)-c(1-v)-1)(T_A-T_B)-\alpha bv T_A + cv(1-T_B) \\
  & \beta_3 = \alpha bvT_A - cv(1-T_B) + (T_A-T_B) \\
  & \beta_4 = 0
\end{split}
\end{equation}
Since $f(\hat{p})$ is a cubic polynomial, three equilibria exist.
We already know that $\hat{p}^*_0 = 0 $ and $\hat{p}^*_1 = 1$ are equilibria.
The third equilibrium can be found by solving $f(\hat{p})/\big(\hat{p}(1-\hat{p})\big)=0$, and it is
\begin{equation} \label{eq:oblique_and_vertic_result}
  \hat{p}^*_2 =  \frac{\alpha bvT_A + cvT_B + T_A-T_B-cv}{c(1-v)\cdot(T_A-T_B) -b (1-\alpha v) \cdot(T_A-T_B)}
\end{equation}

We would like to know when $\hat{p}^*_2$ is legitimate equilibrium, i.e. $0<\hat{p}^*_2<1$.  
To do that we find the derivative  of $f(\hat{p})$,
\begin{equation} \label{eq:derivative_of_phattag-phat2}
  f'(\hat{p}) = 3\beta_0 \hat{p}^2 + 2\beta_1\hat{p}^2 + \beta_3
 \end{equation}
We assume that $T_A>T_B$ and we get that $\beta_1<0$ since $c(1-v)-b(1-\alpha v)<0$ and $c<b$ $v<1$. % why?
In this case the cubic $f(\hat{p})$ goes to infinity when $\hat{p}$ goes to infinity. 
Therefore, if $\hat{p}^*_2$ is positive then $f'(0)$ must be negative % are you sure? I think it must be positive, because i think that the two extremes of a cubic must go in opposite directions
\begin{equation} \label{eq:derivative_of_phattag-phat_p=0}
  f'(0) = \beta_3
 \end{equation}
$\hat{p}^*_2>0$ when $\beta_3<0$.
In addition, $\hat{p}^*_0=0$ is stable when $|f'(0)|<1$ therefore, when $-1<\beta_3<1$.
We can show that
\begin{equation} \label{eq:beta3-condition}
  \beta_3 = \alpha bvT_A - cv(1-T_B) + (T_A-T_B) > -cv(1-T_B) > -1
\end{equation}
Therefore, when $\hat{p}^*_2>0$ we have $-1<\beta_3<0$ and $\hat{p}^*_0=0$ is locally stable.

Now we would like to find when $\hat{p}^*_2<1$.
If $T_A>T_B$, as before, then the cubic $f(\hat{p})$ goes to negative infinity when $\hat{p}$ goes to infinity.
Hence, if $\hat{p}^*_2<1$ then $f'(1)$ must be negative % again, i think that the two extremes of a cubic must go in opposite directions
\begin{equation} \label{eq:derivative_of_phattag-phat_p=1}
  f'(0) = \beta_1 - \beta_3
\end{equation}
Therefore, $\hat{p}^*_2<1$ when $\beta_1-\beta_3<0$.
In addition, $\hat{p}_1=1$ is stable when $|f'(1)|<1$ therefore, when $-1<\beta_1-\beta_3<1$. 
We show that when $0<\hat{p}^*_2<1$ hence $\beta_3<0$ and $\beta_1-\beta_3<0$ we get that $-1<\beta_1-\beta_3$: % cut down sentence to several sentences, it is too complex like this
\begin{equation} \label{eq:beta1-beta3-condition}
 \beta_1 - \beta_3 > - \beta_3 > 0 > -1 ,
\end{equation}
which means that when $0<\hat{p}^*_2<1$ then $\hat{p}^*_1=1$ is locally stable.

\textbf{Result 2.3}: if $\hat{p}^*_2$ is a legitimate equilibrium, then it is locally unstable.

\textbf{Result 2.4}: if $T_A>T_B$, then $\hat{p}^*_2$ is a legitimate equilibrium when $\beta_1<\beta_3<0$.

\textbf{Result 2.5}: if $T_B>T_A$, then $\hat{p}^*_2$ is a legitimate equilibrium when $\beta_1>\beta_3>0$.
